---
title: "Acute Bronchitis Cost Analysis"
format: html
df-print: kable
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

This report primarily investigates whether the total medication cost for acute bronchitis differs between males and females. It also includes a subanalysis of the average medication cost, the distribution of healthcare expenses across income, and the average base cost.

## Load packages

```{r}
library(dplyr)
library(ggplot2)
library(janitor)
library(tidyverse)
library(data.table)
library(zip)
library(rvest)
library(fs)
library(knitr)
library(gt)
library(curl)
```

## Data preparation

Reading in the data and performing data linking and data cleaning.

```{r}
fs::dir_create("data")

if (!fs::file_exists("data/data.zip")) {
  curl::curl_download(
    "https://github.com/STA220/cs/raw/refs/heads/main/data.zip",
    "data/data.zip",
    quiet = FALSE
  )
}

if (!fs::dir_exists("data/data-fixed")) {
  zip::unzip("data/data.zip", exdir = "data")
}

pat_data <- fread("data/data-fixed/patients.csv")
con_data <- fread("data/data-fixed/conditions.csv")
med_data <- fread("data/data-fixed/medications.csv")

patient <- pat_data |>
  select(id, gender, healthcare_expenses, healthcare_coverage, income) |>
  drop_na()

conditions <- con_data |>
  select(start, stop, patient, code, description) |>
  drop_na()

medications <- med_data |>
  select(reasoncode, start, stop, patient, description, base_cost, payer_coverage, dispenses, totalcost, reasondescription) |>
  drop_na()

conditions <- conditions |>
  rename(
    start_con = start,
    stop_con = stop,
    description_con = description
  )

medications <- medications |>
  rename(
    start_med = start,
    stop_med = stop,
    description_med = description
  )

conditions$start_con <- as.Date(conditions$start_con)
conditions$stop_con <- as.Date(conditions$stop_con)
medications$start_med <- as.Date(medications$start_med)
medications$stop_med <- as.Date(medications$stop_med)

conditions <- conditions |>
  filter(code == 10509002)

medications <- medications |>
  filter(reasoncode == 10509002)

patient_conditions <- conditions |>
  left_join(patient, by = c("patient" = "id"))

data_linked <- patient_conditions |>
  inner_join(medications, by = "patient") |>
  filter(
    code == reasoncode,
    start_med == start_con,
    stop_med == stop_con
  )

data_linked$income <- as.numeric(data_linked$income)
data_linked$gender <- factor(data_linked$gender, levels = c("M", "F"))

fwrite(data_linked, "data/data_linked.csv")
```

## Average medication cost

What is the average medication cost for all individuals with the diagnosis, and what is the total medication cost?

```{r}
summary_table_totalcost <- data_linked |>
  summarise(
    `Sum of Total Cost` = round(sum(totalcost, na.rm = TRUE), 2),
    `Average Total Cost` = round(mean(totalcost, na.rm = TRUE), 2)
  )

kable(summary_table_totalcost, caption = "Summary of total medication cost")
```

## Total medication cost by gender

How much does the total medication cost differ between genders?

```{r}
ggplot(data_linked, aes(x = gender, y = totalcost)) +
  geom_boxplot() +
  labs(
    title = "Total Medication Cost by Gender",
    x = "Gender",
    y = "Total Medication Cost"
  ) +
  theme_minimal()
```

## Healthcare expenses across income

What is the distribution of healthcare expenses across income?

The figure below shows the relationship between income and healthcare expenses.

```{r}
ggplot(data_linked, aes(x = income, y = healthcare_expenses)) +
  geom_point() +
  labs(
    x = "Income",
    y = "Healthcare Expenses",
    title = "Healthcare Expenses vs Income"
  ) +
  theme_minimal()
```

## Distribution of healthcare expenses

```{r}
ggplot(data_linked, aes(x = healthcare_expenses)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Healthcare Expenses for Individuals with Acute Bronchitis",
    x = "Healthcare Expenses",
    y = "Count"
  ) +
  theme_minimal()
```

## Distribution of income

```{r}
ggplot(data_linked, aes(x = income)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Income",
    x = "Income",
    y = "Count"
  ) +
  theme_minimal()
```

## Boxplots

```{r}
ggplot(data_linked, aes(y = healthcare_expenses)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Healthcare Expenses",
    y = "Healthcare Expenses"
  ) +
  theme_minimal()
```

```{r}
ggplot(data_linked, aes(y = income)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Income",
    y = "Income"
  ) +
  theme_minimal()
```

## Interpretation

We can observe that both income and healthcare expenses are strongly right skewed. This suggests the presence of high value observations that affect the overall distribution. Income may influence healthcare expenses because differences in income can affect access to care and medication choices. Healthcare expenses may also contribute directly to the total cost of the disease.

Overall, the results suggest that income and healthcare expenses may influence the total medication cost for acute bronchitis.